ANCOVA Comparison Simulations: Imbalanced Treatment Assignment

### 2017-10-05 simulations/imbalanced_ntreat.Rmd
Imbalanced treatment simulations

---
title: "ANCOVA Comparison Simulations: Discrete Outcomes"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Discrete data

Step 1:
We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

@@

```{r generate_sample}
gen_y <- function(gamma, v, error, Z){
  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
  return(y)
}

gen_x <- function(gamma, v, error){
  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
  return(x)
}

generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = multiplier for the magnitude of the treatment effect
  # effect = "same effect" or "heterogeneous"
  # errors = "normal" or "heavy"
  # n = number of individuals at each stratum
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)

@@

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_x(gamma, v, epsilon)
  Y1 <- gen_y(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

plot_power_gamma <- function(powermat_list, gamma_vec, alpha, title){
  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
  melt(gamma_power_alpha, value.name="Power") %>%
  mutate("Method" = Var2) %>%
  mutate("Effect" = rep(gamma_vec, 5)) %>% 
  ggplot(aes(x = Effect, y = Power)) +
  geom_line(aes(color = Method))+
  xlab("Effect size") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}

@@

```{r vary_effect_size_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
The differenced permutation test had substantially lower power than the rest,
again because the correlation between baseline $X$ and outcome $Y$ was low.
In this case, the unadjusted stratified test had comparable power to the linear model based tests.

@@

```{r vary_effect_size_het_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 0.95 ) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
    ylim(0, 0.95) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")

tmp <- ggplot_gtable(ggplot_build(p1))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
mylegend <- tmp$grobs[[leg]]

p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
                         p2 + theme(legend.position="none"),
                         nrow=1),
             mylegend, nrow=2,heights=c(10, 2))

p3
ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)

>>>>

---
title: "ANCOVA Comparison Simulations: Imbalanced Treatment Assignment"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Imbalanced Treatment Groups

Step 1:
We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
Assume that there are $n_j=16$ individuals per stratum.
In this set of simulations, we vary the number of treated units in each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

@@

```{r generate_sample}
gen_y <- function(gamma, v, error, Z){
  y <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
  return(y)
}

gen_x <- function(gamma, v, error){
  x <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
  return(x)
}

generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16), ntreat = c(8, 8, 8)){
  # Input:
  # gamma = multiplier for the magnitude of the treatment effect
  # effect = "same effect" or "heterogeneous"
  # errors = "normal" or "heavy"
  # n = number of individuals at each stratum
  # ntreat = number of treated individuals in each stratum
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)

@@

  # Generate covariates
  Z <- rep(0, N)
  for(i in 1:3){
    Z[stratumID == i] <- rep(1:0, each=c(ntreat[i], n[i] - ntreat[i]))
  }
  Y0 <- gen_x(gamma, v, epsilon)
  Y1 <- gen_y(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

plot_power_ntreat <- function(powermat_list, ntreat, alpha, title){
  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
  melt(gamma_power_alpha, value.name="Power") %>%
  mutate("Method" = Var2) %>%
  mutate("ntreat" = rep(ntreat, 5)) %>% 
  ggplot(aes(x = ntreat, y = Power)) +
  geom_point(aes(color = Method))+
  xlab("Number treated") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}

@@

```{r vary_ntreat_plot, fig.height=4}
load("imbalanced_tr/vary_ntreat_results.Rda")
ntreat_vec <- list( c(8, 8, 8),
                 c(4, 4, 4),
                 c(4, 4, 8),
                 c(4, 8, 8),
                 c(4, 8, 12),
                 c(8, 8, 12),
                 c(8, 12, 12),
                 c(12, 12, 12))
ntreat_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
ntreat_str <- sapply(ntreat_vec, paste, collapse = ",")
ntreat_tot <- sapply(ntreat_vec, sum)
plot_power_ntreat(ntreat_power, ntreat_tot, 0.05, "xx")
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied the number of treated units in each group.
In the random populations that were generated, this corresponded to population average treatment effects of about $1.36$.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% plotted against the total number of treated units.
Table 1 shows the same information.
Each of the tests tended to have higher power when the total number of treated units was around half of the population (in this case, 24 out of 48).
It is interesting to note that all tests had more power when 25\% of units were treated than when 75% were treated; it seems that power is not a symmetric function.

The differenced permutation test had substantially lower power than the rest,
again because the correlation between baseline $X$ and outcome $Y$ was low.
In this case, the unadjusted stratified test had comparable power to the linear model based tests.
ANCOVA had slightly higher power than the linear model permutation tests for all experiments except for when the number of treated units was equal to number of controls and when the number of treated units was 30 or above.

```{r vary_ntreat_tab, results="asis"}
summary05 <- t(sapply(ntreat_power, function(x) x[5,]))
summary05 <- cbind(ntreat_str, summary05)
colnames(summary05) <-  c("Number Treated", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
print(xtable(summary05[order(ntreat_tot), ],
             caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects"),
      include.rownames=FALSE)

@@

```{r vary_ntreat_het_plot, fig.height=4}
load("imbalanced_tr/vary_ntreat_het_results.Rda")
ntreat_het_vec <- list( c(8, 8, 8),
                 c(4, 4, 4),
                 c(4, 4, 8),
                 c(4, 8, 4),
                 c(8, 4, 4),
                 c(4, 8, 8),
                 c(8, 4, 8),
                 c(8, 8, 4),
                 c(4, 8, 12),
                 c(12, 8, 4),
                 c(4, 12, 8),
                 c(8, 12, 8),
                 c(4, 12, 4),
                 c(12, 12, 12))
ntreat_het_power <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
ntreat_het_str <- sapply(ntreat_het_vec, paste, collapse = ",")
ntreat_het_tot <- sapply(ntreat_het_vec, sum)
plot_power_ntreat(ntreat_het_power, ntreat_het_tot, 0.05, "xx")
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied the number of treated units in each stratum and the population average treatment effects were around $1.36$.
Figure 2 and Table 2 show the empirical power at level 5\% for these experiments.
The pattern here is similar to Figure 1: power tends to be highest when the number of treated units is around half of the population.
The treatment effect was largest in the third stratum and smallest in the first stratum;
as expected, Table 2 shows that among experiments that fixed the total number of treated units, the experiment with more treatment units in the third stratum tended to have higher power.
These variations are only slight, though.
As before, the differenced permutation test had low power.
The linear model permutation tests, ANCOVA, and stratified permutation test, in that order, had the best power.

```{r vary_ntreat_het_tab, results="asis"}
summary05het <- t(sapply(ntreat_het_power, function(x) x[5,]))
summary05het <- cbind("Number treated" = ntreat_het_str, summary05het)
colnames(summary05het) <-  c("Number Treated", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
print(xtable(summary05het[order(ntreat_het_tot),],
             caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects"),
      include.rownames=FALSE)
